---
title: "Property Values"
author: "S&DS 361"
date: "2024-02-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
d = read.csv('data/branford.csv')
head(d,2)
```
There is one row per property, each with one single family home, and the columns have the following meanings:

- `value`: the assessed value of the property
- `year`: the year the house was built
- `land`: the square feet of the land area of the property
- `living`: the square feet of the living area of the property
- `beds`: the number of bedrooms
- `address`: the address of the property
- `buildings`: the number of buildings on the property (all 1 in this data set)
- `good`: the condition of the property (10-100)
- `style`: the style of the house (Colonial, Ranch, etc.)
- `model`: all "Residential" in this data set
- `grade`: the grade of the property (A+, A, etc)
- `baths`: the number of full bathrooms
- `halfbaths`: the number of half bathrooms
- `heatfuel`: the type of heating fuel (Oil, Gas, etc.)
- `heattype`: the type of heating (Hot Water, Forced Air-Duc, etc.)
- `ac`: the type of air conditioning (Central, None, etc.)
- `nghd`: a numeric code for the neighborhood of the property
- `stories`: the number of stories in the house
- `saleprice`: the most recent sale price of the property
- `saledate`: the date of the most recent sale
- `roof`: the type of roof (usually Asphalt)
- `bathstyle`: the style of the bathroom (Average or Modern)
- `kitchen_style`: the style of the kitchen (Above Avg, Average or Below Avg)
- `long`: the longitude of the property
- `lat`: the latitude of the property
- `miles_to_coastline`: the distance to the coastline in miles
- `miles_to_highway`: the distance to I-95 in miles
- `miles_to_school`: the distance to the nearest school in miles
- `miles_to_hospital`: the distance to the nearest hospital in miles

Properties with more than one building, and properties with something other than a single family home have been removed. 

## Data exploration

In-class exercise. See Ed Discussions

```{r}

```

